Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models

### 2017-11-09 paper.tex
Update on Overleaf.

\title{Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?}

@@

%\affiliation{\affilnum{1}UMR 8504 G{\'e}ographie-cit{\'e}s, Paris, France\\
%\affilnum{2}Laboratoire Ville Mobilit{\'e} Transport, Universit{\'e} Paris-Est, France\\
%\affilnum{3}Centre for Advanced Spatial Analysis, University College London, UK\\
%\affilnum{4}CNRS, UMR 8097 Centre Maurice Halbwachs, Paris, France\\
%\affilnum{5}UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France\\
%\affilnum{6}Institut des Syst{\`e}mes Complexes Paris Ile-de-France, France}

@@

@@

%My coauthors and I are pleased to submit an original research article entitled ``Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?'' for consideration for publication in Environment and Planning B.

@@

Simulation within computer models has been recognised and increasingly used by geographers as an efficient tool to explore geographical processes, hypotheses and predictive scenarios within virtual laboratories \citep{batty1971modelling, batty2007model, carley1999generating, Quesneletal2009}. %\comment{+ Seb \& James Doran in the 1970s??}
 Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of social and ecological systems in space and time. The specificity of geographical models compared to other social sciences' models is that space and spatial interactions are given a prime role, as geographers are driven by an explicit interest in studying the way space influences the outcomes of the model. Indeed, they usually seek to understand and model the role of space in social interactions and environmental processes, in a city for example. We think simulation can become a very good tool to achieve this, provided that models do include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of output variations to the way space is modelled. Unfortunately, the first condition is not always met, and the second is seldom even mentioned. This paper aims to fill a methodological and conceptual gap, which is a systematic testing of the sensitivity of a model's outcomes to its initial spatial conditions. To demonstrate the genericity of our approach, we develop two applications with classic simulation models commonly used as case studies for comparing and aligning simulation models \citep{Axtelletal1996, wilensky2007making}: \citet{schelling1971dynamic}'s model of segregation and \citet{EpsteinAxtell1996}'s Sugarscape model, both defined at the intra-urban level.\\

@@

Another crucial aspect of spatio-temporal complex systems in general, and cities in particular, is their non-ergodicity~\citep{pumain2012urban}, which means that cross-sectional samples in space (the different cities of a country at time t) are not equivalent to samples in time (the same city at time t, t+1, t+2...) to compute statistics such as average. Similarly to what \citet{gell1995quark} calls \emph{frozen accidents} in complex system generally, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system. Therefore, strong spatio-temporal path-dependencies in the trajectory of individual cities and changing social environments over time prohibits the use of ergodic models... which are the main models to be found in geosimulation! \\

@@

%The example of transportation networks is a good illustration, as their spatial shape and hierarchy is strongly influenced by past investment decisions, technical choices, or political decisions sometimes not rational~(\cite{zembri2010new}). Some aggregated indicators will not take into account positions and trajectories of each agent (such as segregation in the Schelling model) but others, as in the case of spatial patterns of accessibility in a system of cities, fully capture the path-dependency and may therefore be highly dependent of the initial spatial configuration. It is not clear for example what shifted the economical and political capital of France from Lyon to Paris in the early Middle Age, some assumptions being the reconfiguration of trade patterns from South to North of Europe and thus an increased centrality for Paris due to its spatial position: the bifurcation induced by socio-economic and political factors took a deep significance with worldwide repercussions until today when magnified by the spatial configuration. \comment[JR]{quelqu'un sait si d'une part c'est vrai et d'autre part si des references backupent ca ? - je sors ca completement du chapeau pour l'instant mais ça me semble intuitivement raisonnable et une bonne illustration.}[(MLT)L'idée de donner un exemple me semble bonne mais je ne suis pas sûre qu'on l'amène de la bonne manière? --> en fait je crois que l'on a pas besoin de ce paragraphe illustratif, ça complique le papier. Qu'en pensez-vous?]  [(FL) : illustrer par les transports c'est une bonne idée mais pas besoin de parler du moyen âge! la faible croissance de Dijon depuis 30 ans peut être une illustration tarte à la crême. mais pas sûr effectivement qu'on ait absolument besoin d'illustration, en tous cas il n'y en a pas dans le § d'avant (bientôt après ;o)]

@@

%In this section, we detail the method developed to analyse the sensitivity of simulation models to initial spatial conditions. 
The general method workflow of our method is illustrated in Figure \ref{fig:method}. In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with various values of its parameters, we introduced a spatial generator which depends on its own meta-parameters and feeds the model $\mu$ at initialisation (lower branch). The resulting configurations can be clustered into qualitative types of spatial patterns, so that the sensitivity analysis relates to the model's parameters, how the density spatial distribution was generated and the pattern of density in order to produce qualitative conclusion about the model and its behaviour. In particular, we want to emphasize that spatial effects only derive not only from grid size or shape effects, but also from the heterogeneity of relevant geographic entities (people, housing, networks, etc.) in geographical space. In the models we produced, the initial spatial configurations  can be either flat or heterogeneous, monocentric or polycentric, based on external databases and on internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[htbp] \begin{center} 
\resizebox{0.9\textwidth}{!}{ 
	\includegraphics{figures/SchemaMeta_1.png}
} \caption{\textbf{General workflow of our method.}} \label{fig:method} \end{center} \end{figure*} %
%%%%%%%%%%%%%%%%%%%%%%

In order to test for the influence of initial spatial conditions on model outputs, we use a systematic method to compare {\bf phase diagrams}. We define a phase diagram as the vector of model outputs considered as a function of model parameters. Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values $\vec{P}$ as the median of the final values of an output indicator $O$ obtained for the replications of the model initialized with $\vec{P}$.\\

%Comparing phase diagrams is useful because it allows us to compare the 
We have as many phase diagrams than we have spatial grids, which makes a qualitative visual comparison not realistic (with around 50 different spatial configurations for each model experiment). A solution is to use systematic quantitative procedures to compare them. To our knowledge there exist no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature. We propose to define the relative distance $d_r$ between two phase diagrams $\vec{\alpha_1}$ and $\vec{\alpha_2}$ as the share of between-diagrams variability relative to their internal variability, given formally in the case of a one-dimensional phase diagram by

@@

\begin{equation}\label{eq:phase-distance}
d_r\left(\vec{\alpha_1},\vec{\alpha_2}\right) = 2 \cdot \frac{d(f_{\vec{\alpha_1}},f_{\vec{\alpha_2}})^2}{Var\left[f_{\vec{\alpha_1}}\right] + Var\left[f_{\vec{\alpha_2}}\right]}
\end{equation}

where $f_{\vec{\alpha_i}}$ are the phase diagrams at given meta-parameters $\vec{\alpha_i}$. $d$ is a functional distance that we take simply as Euclidian distance. The variances are estimated within each phase diagram $f_{\vec{\alpha_i}}$. For a multi-dimensional phase diagram, we sum these relative distances over the components.

The last methodological point which we need to emphasize is the relation between the workflow we introduce and model exploration workflows. The ideas of multi-modeling and extensive model exploration are nothing from new as ~\cite{openshaw1983data} already advocated for ``model-crunching'', but their effective use only begins to emerge thanks to the apparition of new methods and tools together with an explosion of computation capabilities. The model exploration platform OpenMole~\citep{reuillon2013openmole} allows to embed any model as a blackbox, write modulable exploration workflow using advanced methodologies such as genetic algorithms and distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids. In our case, the workflow tool is a powerful way to embed both the sensitivity analysis and the meta-sensitivity analysis, and allow to couple any generator with any model in a straightforward way as soon as the model can take its spatial configuration as input or from an input file. In this paper, we use the OpenMole platform for spatial environment and model coupling, placing ourselves in the renewed framework of multi-modelling claimed by \citet{cottineau2015modular}.
%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spatial generator of density grids}

The spatial generator applies an urban morphogenesis model \citep{Batty2007}, which has been generalised, explored and calibrated by~\citet{2017arXiv170806743R}. An open implementation and a characterisation of urban forms the model can produce are provided, what allows us to integrate it easily in our workflow. In a nutshell, grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$. More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to patch $i$ with population $P_i$, all $N$ units being added independently and in parallel. At the end of each time step, this growth process is smoothed $n_d$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its extended neighborhood (8 surrounding patches). To avoid border effects such as a reflexion on the border of the world, border patches diffuse to the outside. The procedure stops when a fixed number of steps $t_f$ is reached. The grid then has a population of $t_f \cdot N$ (the population lost due to diffusion process to the outside is reallocated through a normalization procedure at the end of the steps). Grids are thus generated from the combination of the values of these four meta-parameters $\alpha$, $\beta$, $n_d$ and $N$, in addition to the random seed. To ease our exploration, only the distribution of density is allowed to vary rather than the size of the grid, which we fix to a 50x50 square environment. We furthermore fix the total population at $t_f\cdot N = 100,000$, and determine therein the number of steps needed at a given $N$. Typical value ranges for the  parameters will be taken as, following \citet{2017arXiv170806743R}, $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n_d\in\left[1,4\right]$. We illustrate in Fig.\ref{fig:spatialGen} the variety of spatial configuration that can be generated.

%\comment[MLT]{Il me semble que l'on pourrait mieux expliciter la relation N/n (et peut-être du coup inverser N et n). Plus il me semble que l'on devrait dire plus clairement à quoi se réfère alpha : à une cellule $i$? au pas de temps $t$? Pourquoi ne pas indexer alpha? Enfin, je crois que l'on gagnerait encore plus en clarté si on expliquait formellement comment tous les paramètres sont reliés entre eux, notamment pour expliquer l'arrêt de la procédure d'attribution d'une nouvelle pop à la grille.} \comment[FL]{je crois que N c'est l'incrément de pop et n le nombre de fois où c'est smoothé mais c'est quoi la pop totale?}

@@

@@

In order to generate density grids which correspond to empirical density distributions, we select among the generated grids using an objective function which matches the point cloud of 110 metropolitan areas in Europe described by four dimensions of spatial structure : their concentration index, hierarchy index, centrality index and continuity index (cf. \cite{LeNechet2015}). We sample the meta-parameter space using a Latin Hypercube Sampling, which is a convenient technique to have a point cloud with high discrepancy. We sample 2000 points in the 4-dimensional space of parameters {$\alpha$, $\beta$, $n$, $N$}. It yields a subset of 170 interesting grids matching empirical densities, which constituted our set of different initial spatial conditions. These are further clustered into three classes of morphology: compact (e.g. Vienna), polycentric (Liege) and discontinuous (Augsburg) in order to evaluate the non-trivial effects of urban form on simulation results. We select 15 grids of each type to work with in the computation of sensitivity analysis, obtaining a total of 45 spatial configurations to compare (as we will see, the significant number of model runs needed to establish each phase diagram imposes a computational limit on this number of configurations).

%\comment[FL]{pourquoi 15?} \comment[FL]{dire explicitement ce qu'on va faire avec ces catégories dans la partie résultats?}

In the following section, we briefly recall the main components of the two ``classical'' agent-based simulation models used to test how spatial density variations may impact simulation models behaviour and results, and how general the method proposed is.

@@

Schelling's model consists in an abstract urban housing market where agents of different nature sense their environment, evaluate their satisfaction in terms of neighbourhood composition, and relocate if unsatisfied. It has been shown by \cite{Schelling1969} that even tolerant agents tended to produce segregated patterns because of the complexity of their local interactions. The main parameters of this model are the tolerance level (\% of agents similar to {\it ego}), the scope of sensing and the percentage of vacant spaces in the housing market. In addition, we are interested in testing the impact of the spatial distribution of housing capacity in this project, using the generated grids. The outcome of the model is measured as a phase diagram of segregation level, described by the combination of several indicators: Dissimilarity, Moran's Index, Entropy~\citep{brown2006spatial}.

Sugarscape is a model of resource extraction which simulates the unequal distribution of wealth within a heterogenous population \citep{EpsteinAxtell1996}. Although it {\it "is designed to study the interaction of many plausible social mechanisms"} \citep[p.125]{Axtelletal1996}, we refer in this paper to the first (and simplest) version of the model, where {\it "processes allow its agents to look for, move to, and eat a resource ("sugar") which grows on its toroidal array of cells"}. Agents of different vision scopes and different metabolisms harvest a self-regenerating resource available heterogeneously in the initial landscape, they settle and collect this resource, which leads some of them to survive and others to perish. The main parameters of this model are the number of agents, their minimal and maximal resource. In addition, we are interested in testing the impact of the spatial distribution of the resource in this project, using the generated grids. We extend the implementation with agents wealth distribution of~\citet{li2009netlogo}. The outcome of the model is measured as a phase diagram of an index of inequality for resource distribution (Gini index).

\subsection{Experiment design}
For Schelling's model, we explore a grid of a basic parameter space of the model, which is composed of two dimensions: the minimum proportion of similar agents required in the neighbourhood for the agent to be satisfied (or intolerance level) $S\in \left[0;1\right]$ and the vacancy rate of the city $V\in \left[0;1\right]$, which then defines the number of green and red agents, present in equal proportion for this experiment. We sample 1000 parameter values using a Sobol sampling and run 100 repetitions for each configuration. The initial spatial configuration varies across 45 different grids, chosen among the ones which are most representative of the three types of urban morphology: 15 compact grids (like Vienna, cf figure \ref{fig:densityTypes}), 15 polycentric grids (like Liege) and 15 discontinuous grids (like Augsburg). The full experiment thus equates to 4,500,000 simulations (1000 parameter combinations x 45 density grids x 100 replications). We use OpenMOLE to distribute the computation, and apply segregation measures to characterise the results.

For Sugarscape, we explore a grid of a basic parameter space of the model, which is composed of three dimensions: the population of agents $P\in \left[10;510\right]$, the minimal initial agent resource $s_{-}\in \left[10;100\right]$ and the maximal initial agent resource $s_{+}\in \left[110;200\right]$. Each parameter is binned into 10 values, giving 1000 parameter points. We run 50 repetitions for each configuration, what yield reasonable convergence properties. The initial spatial configuration varies across 50 different grids, generated by sampling meta-parameters for the generator in a LHS. We did not use the clustered grids to test the flexibility of our framework, which is demonstrated in this case by a direct sequential coupling of the generator and the model. Indeed, because the density distribution refers to the distribution of resource rather than to the representation of a city structure, we do not need the typology of urban density in this experiment. The full experiment thus equates to 2,500,000 simulations (1000 parameter combinations x 50 density grids x 50 replications).

We choose different design of experiments, both for meta-parameters and for the phase diagram, to demonstrate the robustness of the method to technical choices. In principle, our workflow applies with any manner to generate a spatial configuration (even taking real configurations) and any manner to establish phase diagrams.

@@

\paragraph{Sugarscape} We measure the distance of all phase diagrams to the reference phase diagram computed on the default model setup, which is a bi-centric symmetric configuration (see Fig.~\ref{fig:sugarscape-distance-meta} for its morphological positioning regarded generated grids), using the relative measure defined in equation~\ref{eq:phase-distance}. Indeed, it gives in that case the average squared distance between corresponding points of the phase diagrams, relative to the average of the variance of each. Therefore, values greater than 1 will mean that inter-diagram variability is more important than intra-diagram variability.

@@

We obtain a very strong sensitivity to initial spatial conditions. Indeed, the relative distance of phase diagrams of different density grids to the reference case ranges from 0.09 to 2.98 with a median of 1.52 and an average of 1.30. The mean distance above 1 means that, on average, the model is more sensitive to meta-parameters of grid generation than to its own parameters (population and sugar). Moreover, the maximum distance of 2.98 means that the variation due to the change of grid can be 3 times bigger than the variation due to the model's parameters. We plot in Fig.~\ref{fig:sugarscape-distance-meta} the distribution of these distances in a morphological space. Each point represent one of the 50 different density grids used to initialised the distribution of sugar in the model. The points are projected with respect to the meta-parameters used to generate the grid, and coloured according to the relative distance of the phase diagram of the simulations using this grid to the phase diagram of the reference case. Therefore, the figure ~\ref{fig:sugarscape-distance-meta} shows that the grids generated with a high $\alpha$ (i.e. with a small number of very high density cells) produce simulation results more different compared to the reference case than within this reference case because of parameter variations. This pattern is emphasized when grids are generated with a high $\alpha$ AND a high $\beta$ (i.e. with low gradient of density decrease around the kernels of high density). These grids have the highest relative distance to the reference case. On the contrary, with grids closer to random and uniform patterns (bottom left of the graph), the model's parameters are more important in determining the simulation results.

@@

We consider that the method presented holds great potential for strengthening geographical models' validation. However, two limits and two areas of opportunities have still not been tackled. We discuss them in this section.

@@

As~\citet{pumain2003approche} points out in an overview of complexity approaches in geography, transfer of models and concepts between disciplines may induce a transfer of corresponding assumptions, that become senseless taken out of context. Geography and social science in general have been in the last decades strongly influenced by insights from physics, that beside their enriching

@@

>>>>

\title{Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models
%Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?
}

@@

%\affiliation{\affilnum{1}UMR 8504 G{\'e}ographie-cit{\'e}s, Paris, France\\
%\affilnum{2}Laboratoire Ville Mobilit{\'e} Transport, Universit{\'e} Paris-Est, France\\
%NB : mon affiliation "officielle" c'est : 
%Le Néchet, F., Université Paris-Est, Laboratoire Ville Mobilité Transport, UPEMLV : 5 boulevard Copernic, Cité Descartes F 77454 Marne-la-Vallée cedex 2 France
%\affilnum{3}Centre for Advanced Spatial Analysis, University College London, UK\\
%\affilnum{4}CNRS, UMR 8097 Centre Maurice Halbwachs, Paris, France\\
%\affilnum{5}UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France\\
%\affilnum{6}Institut des Syst{\`e}mes Complexes Paris Ile-de-France, France}

@@

% Dr. Marion Le Texier is associate professor at University of Rouen Normandie. She holds a PhD in geography from both Paris Diderot University and Luxembourg University. Her thesis aimed at modelling the circulation of euro coins through European borders with statistical and agent-based models. Previous to joining Rouen Normandie University in 2016, she has been a Jean Monet fellow at the European University Institute in Florence and a post-doc at Luxembourg University. Her work deals with data and modelling issues in different thematic contexts: access to nature in cities, international online media coverage of earthquakes, to name but two.

% Dr. Florent Le Néchet is associate professor at  Paris-Est Marne-la-Vallée University (UPEM). He holds a PhD in planning from Paris-Est University (UPE). His thesis aimed at modelling the relationships between urban form and mobility patterns at metropolitan level, comparing several European cities, especially in France and Germany. His work deals with charactering and measuring mobility in cities, in relation to the built environment, and involves statistical techniques as well as multi-agent modelling.

% Dr. Romain Reuillon is a research fellow at UMR CNRS 8504 Géographie-cités and at the Complex System Institute Paris Ile-de-France. He holds a PhD in Computer Science from Université Clermont II. His work deals with intensive computation model exploration, focusing on new techniques and methods to extract knowledge from models of simulation. He is the creator of the OpenMole platform for model exploration.

@@

%My coauthors and I are pleased to submit an original research article entitled ``Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models'' for consideration for publication in Environment and Planning B.

@@

Simulation within computer models has been recognised and increasingly used by geographers as an efficient tool to explore geographical processes, hypotheses and predictive scenarios within virtual laboratories \citep{batty1971modelling, batty2007model, carley1999generating, Quesneletal2009}. %\comment{+ Seb \& James Doran in the 1970s??}
 Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of social and ecological systems in space and time. The specificity of geographical models compared to other social sciences' models is that space and spatial interactions are given a prime role, as geographers are driven by an explicit interest in studying the way space influences the outcomes of the model. Indeed, they usually seek to understand and model the role of space in social interactions and environmental processes, in a city for example. We think simulation can become a very good tool to achieve this, provided that models do include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of output variations to the way space is modelled. Unfortunately, the first condition is not always met, and the second is seldom even mentioned. This paper aims to fill a methodological and conceptual gap, which is a systematic testing of the sensitivity of a model's outcomes to its initial spatial conditions. To demonstrate the genericity of our approach, we develop two applications with classic simulation models commonly used as case studies for comparing and aligning simulation models \citep{Axtelletal1996, wilensky2007making}: \citet{schelling1971dynamic}'s model of segregation and \citet{EpsteinAxtell1996}'s Sugarscape model.\\

@@

Another crucial aspect of spatio-temporal complex systems in general, and cities in particular, is their non-ergodicity~\citep{pumain2012urban}, which means that cross-sectional samples in space (the different cities of a country at time t) are not equivalent to samples in time (the same city at time t, t+1, t+2...) to compute statistics such as average. Similarly to what \citet{gell1995quark} calls \emph{frozen accidents} in complex system generally, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system. Therefore, strong spatio-temporal path-dependencies in the trajectory of individual cities and changing social environments over time prohibits the use of ergodic models... which are the main models to be found in geosimulation! \\
% détail de chez détail : j'enlèverais le ! les ... parlent d'eux-mêmes il me semble.

@@

@@

The general method workflow of our method is illustrated in Figure \ref{fig:method}. In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with various values of its parameters, we introduced a spatial generator which depends on its own meta-parameters and feeds the model $\mu$ at initialisation (lower branch). The resulting configurations can be clustered into qualitative types of spatial patterns, so that the sensitivity analysis relates to the model's parameters, how the density spatial distribution was generated and the pattern of density in order to produce qualitative conclusion about the model and its behaviour. In particular, we want to emphasize that spatial effects only derive not only from grid size or shape effects, but also from the heterogeneity of relevant geographic entities (people, housing, networks, etc.) in geographical space. In the models we produced, the initial spatial configurations  can be either flat or heterogeneous, monocentric or polycentric, based on external databases and on internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[htbp] \begin{center} 
\resizebox{0.9\textwidth}{!}{ 
	\includegraphics{figures/SchemaMeta_1.png}
} \caption{\textbf{General workflow of our method.}} \label{fig:method} \end{center} \end{figure*} %
%%%%%%%%%%%%%%%%%%%%%%

In order to test for the influence of initial spatial conditions on model outputs, we use a systematic method to compare \emph{phase diagrams}. We define a phase diagram as the vector of model outputs considered as a function of model parameters. Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values $\vec{P}$ as the median of the final values of an output indicator $O$ obtained for the replications of the model initialized with $\vec{P}$.\\

%Comparing phase diagrams is useful because it allows us to compare the 
We have as many phase diagrams than we have spatial grids, which makes a qualitative visual comparison not realistic (with around 50 different spatial configurations for each model experiment). A solution is to use systematic quantitative procedures to compare them. To our knowledge there exist no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature. We introduce a measure of the relative distance $d_r$ between two phase diagrams $\mu_{\vec{\alpha}_1}$ and $\mu_{\vec{\alpha}_2}$. Phase diagrams are denoted by a common function $\mu$ indexed by what we call \emph{meta-parameters} $\vec{\alpha}$, that capture the spatial configuration (in practice these can be parameters of an upstream model to generate the configuration, or a description of the configuration itself). The $d_r$ measure is taken as the share of between-diagrams variability relative to their internal variability, given formally in the case of a one-dimensional phase diagram by

@@

\begin{equation}\label{eq:phase-distance}
d_r\left(\mu_{\vec{\alpha}_1},\mu_{\vec{\alpha}_2}\right) = 2 \cdot \frac{d(\mu_{\vec{\alpha}_1},\mu_{\vec{\alpha}_2})^2}{Var\left[\mu_{\vec{\alpha}_1}\right] + Var\left[\mu_{\vec{\alpha}_2}\right]}
\end{equation}

where $\mu_{\vec{\alpha}_i}$ are the phase diagrams at given meta-parameters $\vec{\alpha}_i$. $d$ is a functional distance that we take simply as Euclidian distance. The variances are estimated within each phase diagram $f_{\vec{\alpha_i}}$. For a multi-dimensional phase diagram, we sum these relative distances over the components.

The last methodological point which we need to emphasize is the relation between the present workflow and model exploration workflows in general. The ideas of multi-modeling and extensive model exploration are nothing from new as ~\cite{openshaw1983data} already advocated for ``model-crunching'', but their effective use only begins to emerge thanks to the development of new methods and tools together with an explosion of computation capabilities. The model exploration platform OpenMOLE~\citep{reuillon2013openmole} allows to embed any model as a blackbox, to write modulable exploration workflow using advanced methodologies such as genetic algorithms and to distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids. In our case, this tool is a powerful way to embed both the sensitivity analysis and the meta-sensitivity analysis, and to allow the coupling of any spatial generator with any model in a straightforward way as long as the model can take its spatial configuration as an input or from an input file. In this paper, we use the OpenMOLE platform for the spatial environment and the model coupling, placing ourselves in the renewed framework of multi-modelling \citep{cottineau2015modular}.
%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spatial generator of density grids}

The spatial generator applies an urban morphogenesis model \citep{Batty2007}, which has been generalised, explored and calibrated by~\citet{2017arXiv170806743R}. An open implementation and a characterisation of the urban forms which the model can produce are provided, which allows us to integrate it easily in our workflow. In a nutshell, grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$. More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to patch $i$ with population $P_i$, all $N$ units being added independently and in parallel. At the end of each time step, this growth process is smoothed $n_d$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its Moore neighborhood (8 surrounding patches). To avoid border effects such as a reflexion on the border of the world, border patches diffuse to the outside. The procedure stops when a fixed number of steps $t_f$ is reached. The grid then has a population of $t_f \cdot N$ (the population lost due to diffusion process to the outside is reallocated through a normalization procedure at the end of the steps). Grids are thus generated from the combination of the values of these four meta-parameters $\alpha$, $\beta$, $n_d$ and $N$, in addition to the random seed. To ease our exploration, only the distribution of density is allowed to vary rather than the size of the grid, which we fix to a 50x50 square environment. We furthermore fix the total population at $t_f\cdot N = 100,000$, and determine therein the number of steps needed at a given $N$. Typical value ranges for the  parameters will be taken as, following \citet{2017arXiv170806743R}, $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n_d\in\left[1,4\right]$. We illustrate in Fig.\ref{fig:spatialGen} the variety of spatial configurations that can be generated.

@@

In order to generate density grids which correspond to empirical density distributions, we select among the generated grids using an objective function which matches the point cloud of 110 metropolitan areas in Europe described by four dimensions of spatial structure: their concentration index, hierarchy index, centrality index and continuity index (cf. \cite{LeNechet2015}). We sample the meta-parameter space using a Latin Hypercube Sampling, which is a convenient technique to have a point cloud with high discrepancy. We sample 2000 points in the 4-dimensional space of parameters {$\alpha$, $\beta$, $n_d$, $N$}. It yields a subset of 170 interesting grids matching empirical densities, which constituted our set of different initial spatial conditions. These are further clustered into three classes of morphology (figure \ref{fig:densityTypes}): compact (e.g. Vienna), polycentric (Liege) and discontinuous (Augsburg). This clustering allows to evaluate the non-trivial effects of a meaningful urban form on simulation results. We select 15 grids of each type to work with in the computation of sensitivity analysis of an intra-urban model, obtaining a total of 45 spatial configurations to compare (as we will see, the significant number of model runs needed to establish each phase diagram imposes a computational limit on this number of configurations).

@@

In the following section, we briefly recall the main components of the two ``classical'' agent-based simulation models used to test how spatial density variations may impact simulation models behaviour and results, and how general the method is.

@@

Schelling's model consists in an abstract urban housing market where agents of different attributes (for example: red or green) sense their environment, evaluate their satisfaction in terms of neighbourhood composition (how many reds and greens?), and relocate if unsatisfied. It has been shown by \cite{Schelling1969} that even tolerant agents tended to produce segregated patterns because of the complexity of their local interactions. The main parameters of this model are the tolerance level (\% of agents similar to {\it ego}), the scope of sensing, the minority split and the percentage of vacant spaces in the housing market. In addition, we are interested in testing the impact of the spatial distribution of housing capacity in this project, using the generated grids. The outcome of the model is measured as a phase diagram of segregation level, described by the combination of several indicators: Dissimilarity, Moran's Index, Entropy~\citep{brown2006spatial}.

Sugarscape is a model of resource extraction which simulates the unequal distribution of wealth within a heterogenous population \citep{EpsteinAxtell1996}. Although it {\it "is designed to study the interaction of many plausible social mechanisms"} \citep[p.125]{Axtelletal1996}, we refer in this paper to the first (and simplest) version of the model, where {\it "processes allow its agents to look for, move to, and eat a resource ("sugar") which grows on its toroidal array of cells"}. Agents of different vision scopes and different metabolisms harvest a self-regenerating resource available heterogeneously in the initial landscape, they settle and collect this resource, which leads some of them to survive and others to perish. The main parameters of this model are the number of agents, their minimal and maximal resource levels. In addition, we are interested in testing the impact of the spatial distribution of the resource, using the generated grids. We extend the implementation with agents wealth distribution of~\citet{li2009netlogo}. The outcome of the model is measured as a phase diagram of an index of inequality for resource distribution (Gini index).

\subsection{Experiment design}
For Sugarscape, we explore three dimensions of the parameter space: the total population of agents $P\in \left[10;510\right]$, the minimal initial agent resource $s_{-}\in \left[10;100\right]$ and the maximal initial agent resource $s_{+}\in \left[110;200\right]$. Each parameter is binned into 10 values, giving 1000 parameter points. We run 50 repetitions for each configuration, which yields reasonable convergence properties. The initial spatial configuration varies across 50 different grids, generated by sampling meta-parameters for the generator in a LHS. We did not use the clustered grids to test the flexibility of our framework, which is demonstrated in this case by a direct sequential coupling of the generator and the model. Indeed, because the density distribution refers to the distribution of resource rather than to the representation of a city structure, we do not need the typology of urban density in this experiment. The full experiment thus equates to 2,500,000 simulations (1000 parameter combinations x 50 density grids x 50 replications).

For Schelling's model, we also explore three dimensions of the parameter space of the model: the minimum proportion of similar agents required in the neighbourhood for the agent to be satisfied (or intolerance level) $S\in \left[0;1\right]$, the initial split of population, derived from the proportion of green population, $G\in \left[0;1\right]$ and the vacancy rate of the city $V\in \left[0;1\right]$. We sample 1000 parameter values using a Sobol sampling and run 100 repetitions for each configuration. We first try the same experiment design (50 density grids generated on the fly), then look at clustered grids representing urban densities. We choose 45 different grids among the ones which are most representative of the three types of urban morphology: 15 compact grids (like Vienna, cf figure \ref{fig:densityTypes}), 15 polycentric grids (like Liege) and 15 discontinuous grids (like Augsburg). The last experiment thus equates to 4,500,000 simulations (1000 parameter combinations x 45 density grids x 100 replications). We use OpenMOLE to distribute the computation, and apply segregation measures to characterise the results.

We choose different experiment designs, both for meta-parameters and for the phase diagram, to demonstrate the robustness of the method to technical choices. In principle, our workflow applies with any manner to generate a spatial configuration (even taking real configurations) and any manner to establish phase diagrams.

@@

We measure the distance of the phase diagrams for all density grids with respect to the reference phase diagram computed on the default model setup (a bi-centric symmetric configuration) using the relative measure defined in equation~\ref{eq:phase-distance}. For each density grid, we then obtain the average squared distance between corresponding points of the phase diagrams, i.e. the median value of the final output measure, such as segregation or inequality, for a given value of parameters. This average squared distance for each point is then related to the average variance of each of the phase diagram (the reference one and the one for the grid under inquiry). Therefore, values greater than 1 will mean that inter-diagram variability is more important than intra-diagram variability.

\paragraph{Sugarscape}

@@

We obtain a very strong sensitivity to initial spatial conditions for the Sugarscape model. Indeed, the relative distance between the phase diagrams of different density grids and the phase diagram of the reference case ranges from 0.09 to 2.98 with a median of 1.52 and an average value of 1.30. The mean distance above 1 means that, on average, the model is more sensitive to meta-parameters of grid generation than to its own parameters (population and sugar endowment) in the reference model. Moreover, the maximum distance of 2.98 means that the variation due to the change of grid can be up to three times bigger than the variation due to the model parameters. We plot in Fig.~\ref{fig:sugarscape-distance-meta} the distribution of these distances in a morphological space. Each point represent one of the 50 different density grids used to initialise the distribution of sugar in the model. The points are projected with respect to the meta-parameters used to generate the grid, and coloured according to the relative distance of the phase diagram of the simulations using this grid to the phase diagram of the reference case. Therefore, the figure ~\ref{fig:sugarscape-distance-meta} shows that the grids generated with a high $\alpha$ (i.e. with a small number of very high density cells) produce simulation results more different compared to the reference case than within this reference case because of parameter variations. This pattern is emphasized when grids are generated with a high $\alpha$ and a high $\beta$ (i.e. with low gradient of density decrease around the kernels of high density). These grids have the highest relative distance to the reference case. On the contrary, with grids closer to the uniform pattern of the reference case (bottom left of the graph), the model parameters are more important in determining the simulation results.

@@

We consider that the method presented holds great potential for strengthening geographical models' validation. However, two limits and two areas of opportunities have still not been tackled.

@@

As~\citet{pumain2003approche} points out in an overview of complexity approaches in geography, transfer of models and concepts between disciplines may induce a transfer of corresponding assumptions, that become senseless taken out of context. Geography and the social sciences in general have been in the last decades strongly influenced by insights from physics, that beside their highly enriching impact~\citep{o2015physicists}, may have softly imposed strong assumptions such as homogeneity and isotropy of space. We believe that a renewed approach on the role of space as we proposed, in other terms insisting that \emph{space matters}, is an opportunity the for social sciences to build their own stream of methodologies in the modelling domain.

@@

\newpage
